Using an SMT engine to generate Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-14 2017-Xudong-SMT/ActaInformatica.tex
nouveau repertoire ACtaXFIG premiere version Timer-v2 inseree en pages 15-17

\TODO{use a simpler running example (e.g. Lotos Enable operator ?), and move the BIP example to a dedicated section at the end}

As a running example we use the Failure Monitor architecture from the
CubETH nanosatellite on-board software
case-study~\cite{CubETH-case-study} realised using BIP.
The architecture-based design process in BIP takes as input a set of
components providing basic functionality of the system and a set of
temporal properties that must be enforced in the final system.  For
each property, a corresponding architecture is identified
%\,\footnote{%
%
%  The current approach relies on taxonomies of predefined
%  architectures.  Future work may include automatic synthesis of
%  architectures from, \eg CTL~\cite{BaierKatoen2008} formulae.
%
  %}
  and applied to the model, thereby potentially introducing additional
coordinator components and modifying the connectors that define
synchronisation patterns among ports of components.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.8\columnwidth]{XFIG/BIPspec-ArchFailureTimerMax-v2}
  \caption{The BIP specification of the Failure Monitor architecture}
  \label{schema:ArchFailure:BIP}
\end{figure}

Figure~\ref{schema:ArchFailure:BIP} shows a refined version of the
Failure Monitor architecture used in~\cite{CubETH-case-study}.
Contrary to standard BIP models, architectures comprise one or several
\emph{operand} components, whereof only the set of \emph{ports} is
given.  Here, the operand component is $\NameOprnd$ and its interface
consists of the ports $\PortFinish$, $\PortResume$ and $\PortFail$.
The two \emph{coordinator} components\mdash $\NameCtrl$ and
$\NameTimer$\mdash are standard BIP components insofar as they also
have their \emph{behaviour} specified by finite automata extended with
local data variables.  Transitions of these automata are labelled with
the ports of the corresponding components, Boolean guards and update
functions on local variables.  For instance the loop transition
$\tstate{1} \goesto{\PortTick, \GuardTick[], \TimerUpd[]} \tstate{1}$
in the $\NameTimer$ component is labeled by the port $\PortTick$, it
can be fired only when the current value of the local variable $t$ is
greater than $0$.  Upon firing, this transition decrements the value
of $t$ by $1$.  When omitted, the default guard (\resp update
function) is the constant predicate $\true$ (\resp the $\mathit{skip}$
operator).  The constant $\mathrm{Max}$, in $\tstate{0}
\goesto{\PortStart, \TimerInit[]} \tstate{1}$, is a parameter of the
architecture.

Connectors are hierarchical, tree-like structures with component ports
at the leaves.  They define sets of \emph{interactions}, based on the
attributes of the connected ports~\cite{BliSif08-acp-tc}, which may be
either \emph{trigger} (triangles in Fig.~\ref{schema:ArchFailure:BIP})
or \emph{synchron} (bullets in Fig.~\ref{schema:ArchFailure:BIP}).
%
If all sub-connectors of a connector are synchrons, then an
interaction may be executed by the connector only if each subconnector
can contribute.
%
If at least one of the sub-connectors is a trigger, then any
interaction consisting of contributions of any set of sub-connectors,
\emph{involving at least one of the triggers}, can be executed.
%
For instance, the two ports $\NameTimer.\PortStart$ and
$\NameCtrl.\PortFail$ are always synchronised, since they belong to
the same binary sub-connector, where they are both synchrons.  In
particular, this means that whenever the transition $\cstate{0}
\goesto{\PortFail} \cstate{1}$ is fired, so is the transition
$\tstate{0} \goesto{\PortStart, \TimerInit[]} \tstate{1}$,
initialising the timer.  The binary connector $\NameTimer.\PortStart
\bullet\!\!\!-\!\!\!-\!\!\!\bullet \NameCtrl.\PortFail$ is a
sub-connector of a hierarchical connector, where the port
$\NameOprnd.\PortFail$ is a trigger.  Thus, the above interaction can
only happen together with $\NameOprnd.\PortFail$, forming a
ternary interaction.  On the contrary, being a trigger, the port
$\NameOprnd.\PortFail$ can fire alone, forming a singleton
interaction.  The composition semantics of BIP systems consists in
firing exactly one interaction, enabled through at least one of the
top-level connectors, at each execution round.

Finally, \emph{priorities}\mdash defined by a strict partial order on
the set of possible interactions\mdash narrow the choice among the
enabled interactions at any given round.  The default priority is the
so-called \emph{maximal progress}, whereby among any two interactions
$a \subset b$ (as sets of ports), $b$ has higher priority than $a$.
For example, the port $\NameOprnd.\PortFail$ will never fire alone in
a global state, where both $\NameTimer.\PortStart$ and
$\NameCtrl.\PortFail$ are enabled.
%
%% Adding priorities $\NameOprnd.\PortFinish < \NameCtrl.\PortReset$ and
%% $\NameOprnd.\PortFinish < \NameOprnd.\PortResume$, ensures that, after
%% a failure, the operation performed by $\NameOprnd$ cannot finish
%% unless it is explicitly resumed or the system is reset.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{XFIG/FailureTimerPNET}
  \caption{pNet encoding of the Failure Monitor architecture}
  \label{schema:ArchFailure:pNet}
\end{figure}

Figure~\ref{schema:ArchFailure:pNet} shows a pNet encoding of the
above Failure Monitor architecture.  This encoding is structural: each
coordinator component is encoded as a pLTS, the operand
component\mdash as a hole; connectors of the BIP model are encoded as
synchronisation vectors.  Each connector that does not involve
triggers is trivially encoded by a synchronisation vector comprising
the same ports.  In order to encode the semantics of the connectors
involving triggers, we 1)~in the pLTS encoding the coordinator
components, add loop transitions to ensure that all ports involved in
such connectors are enabled in all states, 2)~associate a Boolean
value to each of these ports: the original transitions carry the value
$\true$ (\eg $\cstate{0} \goesto{\PortFail(\true)} \cstate{1}$), the
added loops carry the value $\false$ (\eg $\cstate{2}
\goesto{\PortFail(\false)} \cstate{2}$), 3)~add to the corresponding
synchronisation vectors the Boolean predicate encoding the connector
structure.  For example, $SV0$ encodes the connector discussed above:
the predicate $(b1=b2) \land (b1\lor b2 \Rightarrow b0)$ means that the
``true'' transitions $\NameCtrl.\PortFail$ and $\NameTimer.\PortStart$
can only fire together ($b1 = b2$) and whenever one of them fires,
$\NameOprnd.\PortFail$ must fire also ($b1 \lor b2 \Rightarrow b0$).
This encoding can be systematically obtained for any hierarchical BIP
connector~\cite{BliSif10-causal-fmsd}.
%
Although, for the sake of brevity, we omit priorities from the
encoding, this can be easily achieved, by introducing additional
Boolean variables for relevant ports~\cite{BarBliu15-offer-scico}.

\paragraph{Variations on the FailureTimer example}

\TODO{This is the original example from the Avocs paper, connectors slightly adapted so that it can be composed. Then Max Progress may be required to ensure properties}

For this example, the tool builds 184 open transitions, whereof 173
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 11 open
transitions, is shown in Fig.~\ref{schema:resultOA}.

\begin{figure}[t]
  \centerline{\includegraphics[width=12cm]{XFIG/FailureTimerOA}}
  \caption{Open Automaton for the Failure Monitor architecture}
  \label{schema:resultOA}
\end{figure}

To improve the readability of this figure, we used the following conventions:
we omit the transitions of the two pLTS, and the set of ``working''
holes; and we directly write the resulting action as first element of
each OT, rather than including it as an equality inside the
predicate.
%
%Notice, however that the loop transitions
%$\underline{\mathit{resume}}$ in states $\ostate{s0,t0}$ and
%$\ostate{s2,t0}$, corresponding, respectively, to open transitions
%$ot_2$ and $ot_9$ in Fig.~\ref{schema:ots} of
%App.~\ref{appendix:fullOutput}, involve $\OTvar{resume(false)}$ from
%the two pLTS, \ie only the hole, corresponding to the operand
%component of the Failure Monitor architecture, executes $\PortResume$.

@@

\subsection{Composing 2 FailureTimer BIP Architecture}
\label{section:resultTimerInterval}
\TODO{Enhanced FailureTimer, 2 instances composed, with duplicated control and Timer}

>>>>

As a running example we use the Failure Monitor architecture from the
CubETH nanosatellite on-board software
case-study~\cite{CubETH-case-study} realised using BIP.
The architecture-based design process in BIP takes as input a set of
components providing basic functionality of the system and a set of
temporal properties that must be enforced in the final system.  For
each property, a corresponding architecture is identified
and applied to the model, thereby potentially introducing additional
coordinator components and modifying the connectors that define
synchronisation patterns among ports of components.

\begin{figure}[t]
  \centering
  \includegraphics[width=\columnwidth]{ActaXFIG/BIPspec-ArchFailureTimer-v2-1a}
  \caption{The BIP specification of the Failure Monitor architecture}
  \label{schema:ArchFailure:BIP}
\end{figure}

Figure~\ref{schema:ArchFailure:BIP} shows a refined version of the
Failure Monitor architecture used in~\cite{CubETH-case-study}.
Contrary to standard BIP models, architectures comprise one or several
\emph{operand} components, whereof only the set of \emph{ports} is
given.  Here, the operand component is $\NameOprnd$ and its interface
consists of the ports $\PortResume$ and $\PortFail$.
The two \emph{coordinator} components\mdash $\NameCtrl$ and
$\NameTimer$\mdash are standard BIP components insofar as they also
have their \emph{behaviour} specified by finite automata extended with
local data variables.  Transitions of these automata are labelled with
the ports of the corresponding components, Boolean guards and update
functions on local variables.  For instance the loop transition
$t_1 \goesto{\PortTick, [t < Max], t:=t+1} t_1$
in the $\NameTimer$ component is labeled by the port $\PortTick$, it
can be fired only when the current value of the local variable $t$ is
lower than $Max$.  Upon firing, this transition increments the value
of $t$ by $1$.  When omitted, the default guard (\resp update
function) is the constant predicate $\true$ (\resp the $\mathit{skip}$
operator).  The constant $\mathrm{Max}$ above, is a parameter of the
architecture.

Connectors are hierarchical, tree-like structures with component ports
at the leaves.  They define sets of \emph{interactions}, based on the
attributes of the connected ports~\cite{BliSif08-acp-tc}, which may be
either \emph{trigger} (triangles in Fig.~\ref{schema:ArchFailure:BIP})
or \emph{synchron} (bullets in Fig.~\ref{schema:ArchFailure:BIP}).
%
If all sub-connectors of a connector are synchrons, then an
interaction may be executed by the connector only if each subconnector
can contribute.
%
If at least one of the sub-connectors is a trigger, then any
interaction consisting of contributions of any set of sub-connectors,
\emph{involving at least one of the triggers}, can be executed.
%
For instance, the two ports $\NameTimer.start$ and
$\NameCtrl.\PortFail$ are always synchronised, since they belong to
the same binary sub-connector, where they are both synchrons.  In
particular, this means that whenever the transition 
$c_0 \goesto{\PortFail} c_1$ is fired, so is the transition
$t_0 \goesto{\PortStart, t:=0} t_1$,
initialising the timer.  The binary connector $\NameTimer.\PortStart
\bullet\!\!\!-\!\!\!-\!\!\!\bullet \NameCtrl.\PortFail$ is a
sub-connector of a hierarchical connector, where the port
$\NameOprnd.\PortFail$ is a trigger.  Thus, the above interaction can
only happen together with $\NameOprnd.\PortFail$, forming a
ternary interaction.  On the contrary, being a trigger, the port
$\NameOprnd.\PortFail$ can fire alone, forming a singleton
interaction.  The composition semantics of BIP systems consists in
firing exactly one interaction, enabled through at least one of the
top-level connectors, at each execution round.

%% Finally, \emph{priorities}\mdash defined by a strict partial order on
%% the set of possible interactions\mdash narrow the choice among the
%% enabled interactions at any given round.  The default priority is the
%% so-called \emph{maximal progress}, whereby among any two interactions
%% $a \subset b$ (as sets of ports), $b$ has higher priority than $a$.
%% For example, the port $\NameOprnd.\PortFail$ will never fire alone in
%% a global state, where both $\NameTimer.\PortStart$ and
%% $\NameCtrl.\PortFail$ are enabled.

%
%% Adding priorities $\NameOprnd.\PortFinish < \NameCtrl.\PortReset$ and
%% $\NameOprnd.\PortFinish < \NameOprnd.\PortResume$, ensures that, after
%% a failure, the operation performed by $\NameOprnd$ cannot finish
%% unless it is explicitly resumed or the system is reset.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{ActaXFIG/FailureTimerPNET-v2-1a}
  \caption{pNet encoding of the Failure Monitor architecture}
  \label{schema:ArchFailure:pNet}
\end{figure}

Figure~\ref{schema:ArchFailure:pNet} shows a pNet encoding of the
above Failure Monitor architecture.  This encoding is structural: each
coordinator component is encoded as a pLTS, the operand
component\mdash as a hole; connectors of the BIP model are encoded as
synchronisation vectors.  Each connector that does not involve
triggers is trivially encoded by a synchronisation vector comprising
the same ports.  In order to encode the semantics of the connectors
involving triggers, we 1)~in the pLTS encoding the coordinator
components, add loop transitions to ensure that all ports involved in
such connectors are enabled in all states, 2)~associate a Boolean
value to each of these ports: the original transitions carry the value
$\true$ (\eg $c_0 \goesto{\PortFail(\true)} c_1$), the
added loops carry the value $\false$ (\eg $c_2
\goesto{\PortFail(\false)} c_2$), 3)~add to the corresponding
synchronisation vectors the Boolean predicate encoding the connector
structure.  For example, $SV0$ encodes the connector discussed above:
the predicate $(b1=b2) \land (b1\lor b2 \Rightarrow b0)$ means that the
``true'' transitions $\NameCtrl.\PortFail$ and $\NameTimer.\PortStart$
can only fire together ($b1 = b2$) and whenever one of them fires,
$\NameOprnd.\PortFail$ must fire also ($b1 \lor b2 \Rightarrow b0$).
This encoding can be systematically obtained for any hierarchical BIP
connector~\cite{BliSif10-causal-fmsd}.
%
\noteEM{Decide later if this is still useful}Although, for the sake of brevity, we omit priorities from the
encoding, this can be easily achieved, by introducing additional
Boolean variables for relevant ports~\cite{BarBliu15-offer-scico}.

\paragraph{Computed open automaton}

For this example, the tool (using Algorithm 1) builds 218 open transitions, whereof 206
are detected unsatisfiable by Z3.  The resulting open automaton, with
4 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Fig.~\ref{schema:resultOA1}.

\begin{figure}[t]
  \centerline{\includegraphics[width=12cm]{ActaXFIG/FailureTimerOA-v2-1a}}
  \caption{Open Automaton for the Failure Monitor architecture}
  \label{schema:resultOA1}
\end{figure}

%% To improve the readability of this figure, we used the following conventions:
%% we omit the transitions of the two pLTS, and the set of ``working''
%% holes; and we directly write the resulting action as first element of
%% each OT, rather than including it as an equality inside the
%% predicate.
%
%Notice, however that the loop transitions
%$\underline{\mathit{resume}}$ in states $\ostate{s0,t0}$ and
%$\ostate{s2,t0}$, corresponding, respectively, to open transitions
%$ot_2$ and $ot_9$ in Fig.~\ref{schema:ots} of
%App.~\ref{appendix:fullOutput}, involve $\OTvar{resume(false)}$ from
%the two pLTS, \ie only the hole, corresponding to the operand
%component of the Failure Monitor architecture, executes $\PortResume$.

@@

\subsection{Composing 2 FailureTimer BIP Architecture}
\noteInEM{Modifications stopped here}
\label{section:resultTimerInterval}
\TODO{Enhanced FailureTimer, 2 instances composed, with duplicated control and Timer}
